---
title: "Coursera - Practical Machine Learning - Project"
author: "Oscar Forero"
date: "21 Aug 2014"
output: html_document
---

# Practical Machine Learning - Homework

## Description

Train a machine learning model using the following data set:
[Training](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)

Then use the model to predict the class of the following data set:
[Testing](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)

The data has information from sensors used by people performing a physical exercise, each row is label as correctly performed 'A' or incorrectly perform 'B', 'C', 'D', 'E', the name of the column is 'classe'. 
The model should predict the ‘classe' variable, there is no specific requirement about which other fields to use or not to use.

## Setting up the environment

Import required libraries and set the working directory, you will have to change this if you want to execute the document.

I also use the doMC library to run caret in multiple cores.

```{r, results='hide', message=FALSE}
library(utils)
library(caret)
library(ggplot2)
library(doMC)

setwd("/Users/Oscar/Classes/JohnsHopkins/datasciencecoursera/MachineLearning")
registerDoMC(cores = 4)

```

## Download the data files

The following chunk of code should download
```{r, eval=FALSE}
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", "pml-training.csv", method="internal")

download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", "pml-testing.csv", method="internal")

```

## Loading and cleaning the data, splitting into training and testing data sets

First load the data treating spaces, blanks, divisions by 0 and NA as not a numeric value.
After looking at the data it is clear that many columns are sparsely populated, because there are many other fully populated fields I decided to remove all columns with NA values.

A review of the data shows that some fields are not lectures from the sensors but labeling information and of little relevance for the model creation. For this reason I removed the columns "X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "new_window", "num_window" 

As a final step used the cleaned data set to create a training and a testing data set.

```{r}
inputDS <- read.csv("pml-training.csv", na.strings=c("", "NA", "#DIV/0!"))
naColumns <- apply(inputDS, 2, function(x) any(is.na(x)))

completeDS <- inputDS[,!naColumns]
remove <- c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "new_window", "num_window")

cleanDS <- completeDS[, !names(completeDS) %in% remove]

split <- createDataPartition(cleanDS$classe, p=0.7, list=FALSE)
training <- cleanDS[split,]
testing <- cleanDS[-split,]

summary(training)
```

The predictors left to create the model are:

```{r, echo=FALSE}
featurePlot(x=training[,! names(training) %in% c("classe")], y=training$classe)
```

## A basic model
I decided to use a decision tree model as a benchmark to compare other models of higher complexity.

```{r}
set.seed(6996)

dtModel <- train(classe ~ ., data=training, method="rpart")
dtModel$results
confusionMatrix(dtModel)
```


## Training the model

Next step is to train a model using cross validation to choose the best parameters. 
I decided to try a random forest model first because in my experience it yields good results. 

```{r}
set.seed(90210)

fitControl <- trainControl(method="cv", number=5, repeats=5, classProbs=TRUE)
rfModel <- train(classe ~ ., data=training, method="rf", trControl=fitControl)

rfModel$results
confusionMatrix(rfModel)

```

## Estimated Cross-Validation Error 

```{r}
rfFinal <- rfModel$finalModel
rfFinal

rfOOB <- round(rfFinal$err.rate[rfFinal$ntree, "OOB"]*100, digits=2)
```
The estimated out of bag error from the model as calculated by CARET is `r rfOOB`%. 
It can be found in the finalModel, to extracted I used the formula above which was extracted from the R source code by an anonymous student.

## Error estimation in the Testing Set 

```{r}
pred <- predict(rfModel, newdata=testing)
confMatrix <- confusionMatrix(testing$classe, pred)

accuracy <- confMatrix$overall[1]
error <- round((1 - accuracy)*100, digits=2)

confMatrix
```

In the lectures the recommendation is to still use a testing set that had not been used to train the model.
The error is calculated by subtracting the Accuracy from 1, and in this case is `r error`%


## Conclussion

The Random Forest model takes more time to train than the simple tree model but it has higher accuracy. 
Given that it achieved `r accuracy`% accuracy, Because of that I saw no need to try higher complexity models.

The last step is to use the model to predict the labels for Homework evaluation data set:

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

evalDS <- read.csv("pml-testing.csv")
pred <- predict(rfModel, newdata=evalDS)

pml_write_files(pred)
```

