Health and Economic impact of weather events in USA
========================================================

## Synopsis
* The most damaging to human health in the united states is `r mostDamagingEvent`, which causes `r percentageOfInjuriesForEvent` percent of injuries accross the United States.
* The following `r numberOfEvents` are responsible for `r percentageOfInjures` of the injuries: `r injuriesEvents`
* The most economically damaging event in the united states is `r mostDamagingEvent`, which causes `r percentageOfDamagesPerType` percent of economic damage accross the United States.
* The following `r numberOfDmgEvents` are responsible for `r percentageOfDamagesPerType` of the economic damage: `r damagingEvents`

## Data Processing

### Loading the Raw data from the compress file
This documents requires the R.utils and pylr packages, please intall it before executing the document.
I also used a csv to map the state names from its short to its long form. That csv is also read in here.

```{r cache=TRUE}
library(R.utils)
bunzip2("StormData.csv.bz2",  overwrite=TRUE, remove=FALSE)

rawStorm <- read.csv("StormData.csv")
states <- read.csv("states.csv")

```

### Create tables of Injuries per event, state and event and state 
Grouping Injuries by events and states removing the rows with zero events; because we are looking for the most damaging to the health of the population.

```{r}
library(plyr)

injuriesPerEventAndState <- ddply(rawStorm, .(EVTYPE, STATE), function(x) sum(x$INJURIES))
colnames(injuriesPerEventAndState) <- c("EVTYPE", "STATE", "INJURIES")
injuriesPerEventAndState <- injuriesPerEventAndState[injuriesPerEventAndState$INJURIES > 0, ]
totalInjuries <- sum(injuriesPerEventAndState$INJURIES)
oneP <- totalInjuries / 100

injuriesPerType <- tapply(injuriesPerEventAndState$INJURIES, injuriesPerEventAndState$EVTYPE, sum)
injuriesPerType <- sort(injuriesPerType[injuriesPerType > oneP], decreasing=TRUE)
numberOfInjuries <- sum(injuriesPerType)
percentageOfInjures <- numberOfInjuries * 100 / totalInjuries
injuriesEvents <- rownames(injuriesPerType)
numberOfEvents <- nrow(injuriesPerType)
mostInjurningEvent <- names(injuriesPerType[1])
percentageOfInjuriesForEvent <- injuriesPerType[1] * 100 / totalInjuries

injuriesPerState <- tapply(injuriesPerEventAndState$INJURIES, injuriesPerEventAndState$STATE, sum)
injuriesPerState <- sort(injuriesPerState[injuriesPerState > oneP], decreasing=TRUE)
mostAffectedStates <- names(injuriesPerState)
mostAffectedState <- as.character(states[states$Abbreviation == mostAffectedStates[1], "State"])
numberOfStates <- nrow(injuriesPerState)
mostAffectedStatePercentage <- injuriesPerState[1] * 100 / totalInjuries
```

### Create tables of damage per event, state and event and state 
Grouping damage by events and states removing the rows with zero damage; because we are looking for the most economically damaging ones.

```{r}
DMG <- rawStorm$PROPDMG + rawStorm$CROPDMG
rawStorm$DMG <- DMG
damagePerEventAndState <- ddply(rawStorm, .(EVTYPE, STATE), function(x) sum(x$DMG))
colnames(damagePerEventAndState) <- c("EVTYPE", "STATE", "DMG")
damagePerEventAndState <- damagePerEventAndState[damagePerEventAndState$DMG > 0, ]
totalDamage <- sum(damagePerEventAndState$DMG)
ignoreDmg <- totalDamage / 100 * 2

damagePerType <- tapply(damagePerEventAndState$DMG, damagePerEventAndState$EVTYPE, sum)
damagePerType <- sort(damagePerType[damagePerType > ignoreDmg], decreasing=TRUE)
totalDamagePerType <- sum(damagePerType)
percentageOfDamagesPerType <- totalDamagePerType * 100 / totalDamage
damagingEvents <- rownames(damagePerType)
numberOfDmgEvents <- nrow(damagePerType)
mostDamagingEvent <- names(damagePerType[1])
percentageOfDamageForEvent <- damagePerType[1] * 100 / damagePerType

damagePerState <- tapply(damagePerEventAndState$DMG, damagePerEventAndState$STATE, sum)
damagePerState <- sort(damagePerState[damagePerState > ignoreDmg], decreasing=TRUE)
mostDamagedStates <- names(damagePerState)
mostDamagedState <- mostDamagedStates[1]
numberOfDamagedStates <- nrow(mostDamagedState)
mostDamagedStatePercentage <- damagePerState[1] / totalDamage

```

## Results

### Which type of events are more harmful to the population health
* The most damaging to human health in the united states is `r mostInjurningEvent`, which causes `r percentageOfInjuriesForEvent` percent of injuries accross the United States.
* The following `r numberOfEvents` are responsible for `r percentageOfInjures` of the injuries: `r injuriesEvents`
* `r mostAffectedState` is the most affected state which suffers `r mostAffectedStatePercentage` percent of weather related injuries in the United States.

```{r fig.width=14, fig.height=8}
par(mfrow=c(1,2))

barplot(injuriesPerType / totalInjuries, 
        main="Events causing most injuries", xaxt='n',
        ylab="Proportion of Injuries Caused",
        legend.text=injuriesEvents, col=rainbow(numberOfEvents))
Axis(side=1, labels=FALSE)

plotN = 20
stateNames <- lapply(head(mostAffectedStates,plotN), 
                     FUN=function(x) as.character(states[states$Abbreviation == x, "State"]))
barplot(head(injuriesPerState,plotN) / totalInjuries, 
        main="Most Affected States", xaxt='n',
        ylab="Proportion of Injuries",
        legend.text=stateNames, col=rainbow(plotN))
Axis(side=1, labels=FALSE)
```

### Which type of events cause more economic damage
* The most economically damaging event in the united states is `r mostDamagingEvent`, which causes `r percentageOfDamagesPerType` percent of economic damage accross the United States.
* The following `r numberOfDmgEvents` are responsible for `r percentageOfDamagesPerType` of the economic damage: `r damagingEvents`
* `r mostDamagedState` is the most affected state which suffers `r mostDamagedStatePercentage` percent of weather related economic damage in the United States.

```{r fig.width=14, fig.height=9}
par(mfrow=c(1,2))

plotN = 20
barplot(head(damagePerType,plotN) / totalDamage, 
        main="Most Economically Damaging Events", xaxt='n',
        ylab="Proportion of losses caused",
        legend.text=head(damagingEvents, plotN), col=rainbow(numberOfDmgEvents))
Axis(side=1, labels=FALSE)

plotN = 20
stateNames <- lapply(head(mostDamagedStates,plotN), 
                     FUN=function(x) as.character(states[states$Abbreviation == x, "State"]))
barplot(head(damagePerState,plotN) / totalDamage, 
        main="Most Economically Affected States", xaxt='n',
        ylab="Proportion of losses",
        legend.text=stateNames, col=rainbow(plotN))
Axis(side=1, labels=FALSE)
```
