# Reproducible Research: Peer Assessment 1

This is my solution to the first project. 
I am using a single repository for all the specialization classes instead of cloning a repo for each class.

## Loading and preprocessing the data


```{r}
library(ggplot2)

# Replace this for the right directory 
script.dir <- "/Users/Oscar/Classes/JohnsHopkins/datasciencecoursera/ReproducibleResearch/Project_1"
setwd(script.dir)

activityRaw <- read.csv("activity.csv", colClasses=c(NA, "Date", "character"), stringsAsFactors=FALSE)
activityRaw$interval = sprintf("%04i", as.numeric(activityRaw$interval))
summary(activityRaw)

perDay <- tapply(activityRaw$steps, activityRaw$date, sum, na.rm=TRUE)
days <- names(perDay)
dailyActivity <- data.frame(days, perDay)
colnames(dailyActivity) <- c("date", "steps")
summary(dailyActivity)

perInterval <- tapply(activityRaw$steps, activityRaw$interval, mean, na.rm=TRUE)
intervals <- names(perInterval)
intervalActivity <- data.frame(intervals, perInterval, stringsAsFactors=FALSE)
colnames(intervalActivity) <- c("interval", "steps")
summary(intervalActivity)

avgSteps = mean(dailyActivity$steps, na.rm=T)
medianSteps = median(dailyActivity$steps, na.rm=T)


```

## What is mean total number of steps taken per day?

The average number of steps is: 9354
The median number of steps is: 10395

This is calculate with the following code:
```{r}  
avgSteps = mean(dailyActivity$steps, na.rm=T)
avgSteps

medianSteps = median(dailyActivity$steps, na.rm=T)
medianSteps
```

### Histogram of daily number of steps
```{r fig.width=7, fig.height=6}  

ggplot(dailyActivity, aes(x=steps)) + 
    geom_histogram(binwidth=2000, colour="black", fill="white") +
    xlab("Steps per day") +
    ylab("Number of Days") +
    ggtitle("Frequency of steps per day") +
    geom_vline(aes(xintercept=avgSteps),   
               color="red", linetype="dashed", size=1, show_guide=TRUE) +
    geom_text(mapping=aes(x=avgSteps, y=0, label="Mean Steps per day"), 
              size=6, angle=90, vjust=-0.5, hjust=-0.1) +
    geom_vline(aes(xintercept=medianSteps),   
               color="blue", linetype="dashed", size=1, show_guide=TRUE) +
    geom_text(mapping=aes(x=medianSteps, y=0, label="Median Steps per day"), 
              size=6, angle=90, vjust=1.3, hjust=-0.1) 
    
```

## What is the average daily activity pattern?

### Most active interval
The most active interval starts at 8:35 AM, it can be extracted from the summary per interval as shown in the following code snippet:

```{r}
maxActivityInterval <- intervalActivity[which.max(intervalActivity$steps),1]
maxActivityInterval
```

### Time Series of Averge Steps per Interval
```{r fig.width=7, fig.height=6}
noon = 144
ggplot(intervalActivity, aes(x=interval, y=steps, group=1)) +
    geom_line() +
    geom_vline(aes(xintercept=noon),   
               color="blue", linetype="dashed", size=1, show_guide=TRUE) +
    geom_text(mapping=aes(x=noon, y=0, label="Noon"), 
              size=6, angle=90, vjust=1.3, hjust=-0.1) 
```

## Imputing missing values

2304 rows have missing data, and it can be calculated with the following code
```{r}
missing = complete.cases(activityRaw)==FALSE

# Number of missing using a sum
sum(missing)

```

I choose the strategy of using the average for the 5 minute interval. After imputing the data has a normal distribution with the median equal to the mean.

To achieve that used the following code:
```{r}
imputed = activityRaw

for (interval in intervalActivity$interval) {
  imputed[is.na(imputed$steps) & imputed$interval==interval, c("steps")] = 
    intervalActivity[intervalActivity$interval==interval, c("steps")] 
}
```

With the new data as base then I can create the summary per day, and do the requested graph.
```{r fig.width=7, fig.height=6}

perDay <- tapply(imputed$steps, imputed$date, sum, na.rm=TRUE)
days <- names(perDay)
dailyActivity <- data.frame(days, perDay)
colnames(dailyActivity) <- c("date", "steps")
summary(dailyActivity)

avgSteps = mean(dailyActivity$steps, na.rm=T)
medianSteps = median(dailyActivity$steps, na.rm=T)

print(paste("Averge number of steps per day: ", avgSteps))
print(paste("Median number of steps per day: ", medianSteps))

ggplot(dailyActivity, aes(x=steps)) + 
    geom_histogram(binwidth=2000, colour="black", fill="white") +
    xlab("Steps per day") +
    ylab("Number of Days") +
    ggtitle("Frequency of steps per day") +
    geom_vline(aes(xintercept=avgSteps),   
               color="red", linetype="dashed", size=1, show_guide=TRUE) +
    geom_text(mapping=aes(x=avgSteps, y=0, label="Mean/Median Steps per day"), 
              size=6, angle=90, vjust=-0.5, hjust=-0.1) 
    
```

## Are there differences in activity patterns between weekdays and weekends?

In order to answer this question first we need to add a factor variable to the data frame.

```{r}
weekdays = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
weekends = c("Saturday", "Sunday")
imputed$day = weekdays(imputed$date)

imputed[imputed$day %in% weekdays, c("day")] = "weekday"
imputed[imputed$day %in% weekends, c("day")] = "weekend"
imputed$day = as.factor(imputed$day)

```

### Average number of steps by interval for weekdays/weekends
```{r fig.width=7, fig.height=6}
weekdays <- imputed[imputed$day == "weekday",]
weekend <- imputed[imputed$day == "weekend",]

perIntWeekday <- tapply(weekdays$steps, weekdays$interval, mean)
intervals <- names(perIntWeekday)
intWeekday <- data.frame(intervals, perIntWeekday, stringsAsFactors=FALSE)
intWeekday$day = "weekday"
colnames(intWeekday) <- c("interval", "steps", "day")

perIntWeekend <- tapply(weekend$steps, weekend$interval, mean)
intervals <- names(perIntWeekend)
intWeekend <- data.frame(intervals, perIntWeekend, stringsAsFactors=FALSE)
intWeekend$day = "weekend"
colnames(intWeekend) <- c("interval", "steps", "day")

intervalActivity = rbind(intWeekday, intWeekend)
intervalActivity$day = as.factor(intervalActivity$day)

noon = 144
ggplot(intervalActivity, aes(x=interval, y=steps, group=1)) +
    geom_line() +
    facet_grid(day ~ .) +
    geom_vline(aes(xintercept=noon),   
               color="blue", linetype="dashed", size=1, show_guide=TRUE) +
    geom_text(mapping=aes(x=noon, y=0, label="Noon"), 
              size=6, angle=90, vjust=1.3, hjust=-0.1) 
```
